---
title: "Identifying changes in RNA binding using a linear model"
output:
  pdf_document: default
  html_notebook: default
header-includes:
- \usepackage{xcolor}
- \usepackage{framed}
---


\colorlet{shadecolor}{lightgray!10} 

In this notebook, we will identify changes in RNA binding following parallel quantification of the "total" and "RNA-bound" protein abundance using the OOPS protocol. We start with a single protein and use the base function `lm`, then apply this to the complete dataset to identify all proteins with a change in RNA binding. Following this, we use the `limma` package to test for RNA binding changes using a moderated t-statistic.

Below we load the required packages and set a plotting theme.
```{r, message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
library(biobroom)
library(MSnbase)
library(limma)

# set up standardised plotting scheme
theme_set(theme_bw(base_size = 20) +
            theme(panel.grid.major=element_blank(),
                  panel.grid.minor=element_blank(),
                  aspect.ratio=1))

cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#999999")
```


We start by reading in the data. Our input here is the protein-level quantification for the Nocodazole arrest/release experiment conducted for the OOPS NBT paper (https://www.nature.com/articles/s41587-018-0001-2). In this experiment, we wanted to assess changes in RNA binding in arrested/released cells. To do this, we quantified "total" protein abundance and RNA-bound (extracted by OOPS) protein abundance. The peptide-level abundances have been aggregated to protein level abundance and center-median normalised. Proteins with missing values have been removed. Only proteins quantified in both "total" and "OOPS" samples are included.

The input data here is identical to supplementary table 5 from the above paper. 
```{r}
protein_quant_raw <- readxl::read_excel('./ncbi_30607034_OOPS_NBT_table_S5.xlsx',
                                        sheet=3, skip=1, n_max=1917) %>%
  tibble::column_to_rownames('master_protein')

```

We want to store this data in an MSnSet since this is the data structure one would normally use to store proteomics data within R.

As a reminder, the `MSnSet` class mimics the `ExpressionSet` class and contains 3 matrices:

1. assay data (obtained via: `exprs`)

2. feature data (`fData`)

3. phenotype data (`pData`)

The assay data is the quantification of the features (could be PSMs/peptides/proteins) and contains one column per sample

The feature data describes each feature, e.g peptide sequence, master protein accession, retention time etc etc

The phenotype data describes the samples


Below, we read the abundances into the assay data. Then we take the column names of the assay data (the sample names) and separate out the components of these sample names to provide the descriptions for the samples. Note that we do not have any feature data here since the input data does not decribe any aspects of the features (proteins).
```{r}
protein_quant <- MSnSet(exprs=as.matrix(protein_quant_raw))

pData(protein_quant) <- data.frame(sample_name=colnames(protein_quant_raw)) %>%
  separate(sample_name, into=c('timepoint', 'replicate', 'type'), remove=FALSE) %>%
  tibble::column_to_rownames("sample_name")
```

Let's take a look at our protein quantification data. If we `print` the object, we get a summary including the processing steps performed. 

Here we have 1916 features (proteins) quantified across 6 samples. 

```{r}
print(protein_quant)
```


Here's the top of the assay data.
```{r}
print(head(exprs(protein_quant), 2))
```


And here is the phenotype data. As we can see, we have 3 replicates each and three timepoints. 
```{r}
print(pData(protein_quant))
```

For our purposes, we're only going to focus on the 0h and 6h timepoints phases so we can remove the other data. Below, we subset to the 0h and 6h timepoints
```{r}
protein_quant <- protein_quant[,pData(protein_quant)$timepoint %in% c('0h', '6h')]
```

To run `lm`, we will convert the data format into a "tidy" long format `data.frame` using `biobroom::tidy()`
```{r}

protein_quant_long <- protein_quant %>% 
  tidy(addPheno=TRUE)  %>% # "tidy" the object, e.g make it into a tidy data format --> long
  mutate(intensity=value) %>% dplyr::select(-value) # rename the "value" column -> "intensity"

```

Top of the protein expression `data.frame`. See how each intensity value now has it's own row with the other columns describing the associated aspects of the intensity value, e.g the protein and experimental condition
```{r}
print(head(protein_quant_long))
```


We want to tell R which is the order of the values in the condition and type columns so that the fold changes are in the expected direction, e.g positive = higher in 6h vs 0h.
```{r}
protein_quant_long$timepoint <- factor(protein_quant_long$timepoint, levels=c("0h", "6h"))
protein_quant_long$type <- factor(protein_quant_long$type, levels=c("total", "OOPS"))

```



Now we model the protein intensity according to the models described in `1_simple_example_vd.Rmd`. As an example, let's see the results from just applying the models to a single UniprotID.
```{r, out.width = '70%'}
protein_quant_long %>% filter(protein == 'A0AVT1') %>%
  ggplot(aes(timepoint, intensity, group=1)) +
  geom_point(size=2) +
  stat_summary(geom="line", fun.y=mean) +
  xlab("") +
  facet_wrap(~type)

fit <- protein_quant_long %>% filter(protein == 'A0AVT1') %>%
  lm(formula=intensity~timepoint*type)

print(summary(fit))

```

We can see that the model fits the data well ("Multiple R-squared:  0.9673,	Adjusted R-squared:  0.955 "). We can see that the interaction term that we're interested in for changes in RNA binding ("timepoint6h:typeOOPS") significantly deviates from zero.

Below, we make a function to run the linear model for a single protein
```{r}
Change_in_RNA_binding_LM <- function(obj){
  
  fit <- obj %>% lm(formula=intensity ~ timepoint + type + timepoint*type)
  
  fit_values <- c(coef(summary(fit))["timepoint6h:typeOOPS",],
                  summary(fit)$adj.r.squared)
  
  names(fit_values) <- c("lm_fold_change", "lm_std_error", "lm_t_value",
                         "lm_p_value", "lm_adj_R_squared")
  fit_values <- as.data.frame(t(fit_values), stringsAsFactors=FALSE)
  
  return(fit_values)
}

```

Testing the above function with our example protein
```{r}
protein_quant_long %>% filter(protein == 'A0AVT1') %>% Change_in_RNA_binding_LM()
```

Below, we make a function to run the `Change_in_RNA_binding_LM ()` function on all proteins in turn using dplyr. We will use the standard Benjamini-Hochberg method to adjust p-values for the multiple tests we have conducted.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
runLM <- function(obj){
  results <- obj %>%
    group_by(protein) %>% # group the data frame by the unique protein values
    # Apply the Change_in_RNA_binding_LM functions to each group; "." here is the whole data frame 
    do(Change_in_RNA_binding_LM(.))

  results$lm_BH <- p.adjust(results$lm_p_value, method="BH")# FDR 
  
  return(results)
   
}

lm_results <- protein_quant_long  %>% runLM()

```

Below, we plot the p-values. Under the null hypothesis they should show an approximately uniform distribution. If there were a large number of proteins with a significant change in RNA binding, we would expect an additional "spike" with low p-values (<0.05). We see an approximately uniform distribution with a large 'spike' of low p-values but also a slight skew towards low p-value. This may indicate the presence of changes in RNA binding which we are insufficiently powered to detect.
```{r, out.width = '50%'}
lm_results %>% ggplot(aes(lm_p_value)) + geom_histogram(bins=20)
``` 


We can use a volcano plot to take a look at the estimated fold changes and associated p-values
```{r, out.width = '50%'}


lm_results %>%
  mutate(sig=ifelse(lm_BH<0.01, "Sig.", "Not sig.")) %>% # add "sig" column
  ggplot(aes(x=lm_fold_change, y=-log10(lm_p_value), colour=sig)) +
  geom_point(size=0.25) + 
  scale_colour_manual(values=c("black", cbPalette[6]), name="") + # manually adjust colours
  xlab("Fold change (log2)") + ylab("-log10(p-value)") # manual axes labels

```

Now, let's apply `limma` to performed a moderated version of the above linear modeling. For a full explanation of the steps in a `limma` analysis, see: https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf

First of all we need to create a design matrix. We can do this from the `pData` since this contains the information about the samples

```{r}
timepoint <- factor(protein_quant$timepoint, levels=c("0h", "6h"))
type <- factor(protein_quant$type, levels=c("total", "OOPS"))

design <- model.matrix(~timepoint*type) 

limma_fit_results <- lmFit(protein_quant, design)

limma_fit_results <- contrasts.fit(limma_fit_results, coefficients="timepoint6h:typeOOPS")

limma_fit_results <- eBayes(limma_fit_results, trend=TRUE, robust=TRUE) # Bayesian shrinkage
```

Visualising the relationship between protein abundance and variability
```{r, out.width = '70%'}
plotSA(limma_fit_results)
```

Below, we summarise the number of signficant p-values (post BH FDR correction) using a 1% FDR threshold. 
```{r}
summary(decideTests(limma_fit_results, p.value=0.01, adjust.method="BH"))
```

Below, we extract the limma results
```{r, out.width = '60%'}
limma_results <- topTable(limma_fit_results, coef = "timepoint6h:typeOOPS", n = Inf, confint=TRUE)
```

And then make the volcano plot for the `limma` RNA binding changes
```{r, out.width = '70%'}
limma_results %>%
  mutate(sig=ifelse(adj.P.Val<0.01, "Sig.", "Not sig.")) %>% # add "sig" column
  ggplot(aes(x=logFC, y=-log10(P.Value), colour=sig)) +
  geom_point(size=0.25) + 
  scale_colour_manual(values=c("black", cbPalette[6]), name="") + # manually adjust colours
  xlab("Fold change (log2)") + ylab("-log10(p-value)") # manual axes labels

```


Finally, we can get `limma` to return the signficant changes using both p-value and log-fold change thresholds using the TREAT method to explictly test the null hypothesis that the fold change is less than our specified threshold.  (https://www.ncbi.nlm.nih.gov/pubmed/19176553). 


```{r}
# Test null hypothesis than change is <2-fold 
limma_results_treat <- treat(limma_fit_results,lfc=1) 

# subset to proteins passing the 1% FDR threshold for absolute fold change > 2
limma_results_treat_sig <- topTreat(limma_results_treat, coef = "timepoint6h:typeOOPS", n = Inf,
                                p.value=0.01, lfc=1, adjust.method="fdr", confint=0.95)

```

And below we reproduce our volcano plot including the 95% confidence interval and highlight those proteins which have < 1% FDR and an absolute fold change significantly greater than 2. 

```{r, fig.height=6, fid.width=6, out.width = '60%'}

limma_results$sig <- ifelse(limma_results$P.Value<=0.01, "<1% FDR", ">1% FDR") # add "sig" column
limma_results$sig[
  rownames(limma_results) %in% rownames(limma_results_treat_sig)] <- "<1% FDR & FC > 2 (TREAT)"

limma_results$SE <- sqrt(limma_fit_results$s2.post) * limma_fit_results$stdev.unscaled[,1]

p <- limma_results %>%
  ggplot(aes(logFC, -log10(P.Value), colour=sig)) +
  geom_point(size=1) +
  scale_colour_manual(values=c(cbPalette[c(6,2)], "grey20"), name="") + # manually adjust colours
  geom_vline(xintercept=1, linetype=2, colour="grey70") +
  geom_vline(xintercept=-1, linetype=2, colour="grey70") +
  theme(legend.position="top", legend.direction=2)
  
print(p)
```

We can also add error bars to the above Below, we can see that many of the proteins with a fold change (FC) point estimate > 2 have a 95% confidence interval that overlaps the dashed lines for >2-fold change. TREAT also takes the multiple testing into account so it's even more conservative than just using the 95% CI shown below.
```{r, fig.height=6, fid.width=6, out.width = '60%'}
print(p + geom_errorbarh(aes(xmin=CI.L, xmax=CI.R)))
```

Of course, the threshold for the fold changes you are interested in depends entirely on your prior expectations. A 2-fold threshold for "biologically" relevant changes will likely be overly conservative in the great majority of cases.




